Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-07-27 paper.tex
more work

\section{Dependent Types}
In this section, we introduce the concept of dependent types and their role in
type theoretic systems. At their core, dependent types can be thought of simply
as an \textit{interplay of types and terms}. To understand this relationship,
it helps to think of the ways in which types and terms can \textit{depend} on
one another. Altogether, this results in four dependencies:

@@

\(\Pi\)-types are actually significantly more expressive than the ones in
\(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). In a later system, we
allow \(\Pi\)-types to be parameterized over terms (e.g. {\tt ∀ (x : t) → t}).

@@

In the process of introducing dependent types, we have shown the necessary
changes and extensions on the STLC to achieve the expressivity necessary to
function as a foundation for mathematics. To clarify this process and to show
the relationship between each of the systems we have previously mentioned, we
include a diagram of the {\em Lambda Cube} \cite{}:
\begin{center}
  \begin{tikzpicture}[
      back line/.style={densely dotted},
      cross line/.style={preaction={draw=white, -,
          line width=6pt}}]
    \matrix (m) [matrix of math nodes,
      row sep=1em, column sep=.75em,
      text height=1ex,
      text depth=0.5ex]{
      & \lambda\omega & & \lambda C \\
      \lambda 2 & & \lambda P2 \\
      & \lambda\underline\omega & & \lambda P\underline\omega \\
      \lambda\!\!\rightarrow  & & \lambda P \\
    };
    \path[-]
    (m-1-2) edge (m-1-4)
    edge (m-2-1)
    edge [back line] (m-3-2)
    (m-1-4) edge (m-3-4)
    edge (m-2-3)
    (m-2-1) edge [cross line] (m-2-3)
    edge (m-4-1)
    (m-3-2) edge [back line] (m-3-4)
    edge [back line] (m-4-1)
    (m-4-1) edge (m-4-3)
    (m-3-4) edge (m-4-3)
    (m-2-3) edge [cross line] (m-4-3);
  \end{tikzpicture}
\end{center}
Reading the cube requires us to start on the edge marked \(\lambda\!\!\rightarrow\),
which corresponds to the STLC, the base system. From here, we can either move
inwards, upwards, or rightwards, which corresponds to the extension to the other
systems marked on the respective edges. Using these three movements, we
eventually reach \(\lambda C\), leaving it in the {\em most powerful} position
on the cube.

@@

prop' : {A : ⊥ → Set} → (x : ⊥) → A x
prop' {A} perp = ind⊥ A perp
\end{code}
\end{minipage}
\end{center}
The behavior of the empty type is meant to model {\it ex-falso} (i.e., from
absurdity comes anything we like). In general, we use the empty type when
writing proofs containing negated expressions.

@@

ind⊤ : (C : ⊤ → Set) → C * → (x : ⊤) → C x
ind⊤ C c * = c
\end{code}
\end{minipage}
\end{center}
Similar to \(\perp\), \(\top\) is simple and is rather uninteresting by itself.
There are, however, some trivial yet interesting proofs that involve the two.
For the first of these, we also introduce a special kind of proof: {\it uniqueness
of identity} (UIP) \cite{}. In this proof, we set out to determine and classify
the identity of every element within a given type. For example, for \(\top\), we
should be able to prove:
$$\forall (x : \top) \rightarrow x ≡ *$$
That is, for every element of type \(\top\), that element is equivalent to {\tt *}.
This second expression, {\it x ≡} {\tt *}, is an example of an {\it identity type},
which we describe later on in this section. For now, we simply say that the
sole constructor for the identity type is {\it refl x}, for all x.

Now, to prove the above proposition, we can use the induction principle for
\(\top\), since the proposition we are trying to prove concerns all elements
of \(\top\):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ : (x : ⊤) → x ≡ *
uip⊤ = ind⊤ (λ top → top ≡ *) (refl *)
\end{code}
\end{minipage}
\end{center}
Here, we have taken our motive to be the expression {\tt (λ top → top ≡ *)},
which essentially repeats the information already given by our type expression.
The second element, {\tt refl *}, is a proof that our motive holds in the case that
{\tt top} is {\tt *} and is of type {\tt * ≡ *}.

Alternatively, we could have written our proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ * = refl *
\end{code}
\end{minipage}
\end{center}
This form of proof uses pattern matching and essentially asserts that the only
element in \(\top\) is {\tt *}, which greatly simplifies our proof. Both methods
of proof are correct, but, in general, using pattern matching is only safe when
the values of proof expressions are not dependent one another (i.e., their
values don't change depending on which expression we apply pattern matching to).

@@

\section{More Types and Sample Proofs}
In this section, we present the type of Booleans, Natural numbers, Lists,
and the identity type. To further understand these types, we also include
a brief description of the different forms of equality, namely definitional,
judgmental and propositional. In type theory, the specific differences
between the three are generally misunderstood but are important in understanding
basic type theoretic proof methods, which we also go into some detail in
this section with several examples.

\subsection*{Equality}
%% talk about definitional equality, judgmental equality, prop equality

\subsection*{Booleans}
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data 𝔹 : Set where
  True  : 𝔹
  False : 𝔹

rec𝔹 : (C : Set) → C → C → 𝔹 → C
rec𝔹 C t f True = t
rec𝔹 C t f False = f

ind𝔹 : (C : 𝔹 → Set) → C True → C False → (x : 𝔹) → C x
ind𝔹 C t f True = t
ind𝔹 C t f False = f
\end{code}
\end{minipage}
\end{center}

@@

\subsection*{Identity Types}
%% say something that this is the HoTT version of identity
%% possibly also point out that the discrepancy between the two
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
infix  4  _≡_
data _≡_ {ℓ} {A : Set ℓ} : (a b : A) → Set ℓ where
  refl : (a : A) → (a ≡ a)

-- "PROPOSITIONAL EQUALITY"
-- For a proposition, C, on a proof of x ≡ y for
-- any x and y, if we can show that it holds for
-- (refl x), then C holds for all proofs, p : x ≡ y.
J : {A : Set} (C : {x y : A} → x ≡ y → Set) →
    (∀ (x : A) → C (refl x)) →
    {x y : A} (p : x ≡ y) → C p
J C c (refl x) = c x

-- "STRICT EQUALITY"
-- For a proposition, C, on a reflexive equality (x ≡ x),
-- for any x, if we can show that it holds for (refl x), then
-- C holds for all other reflexive equalities p : x ≡ x.
K : {A : Set} (C : {x : A} → x ≡ x → Set) →
    (∀ (x : A) → C (refl x)) →
    {x : A} (p : x ≡ x) → C p
K C c (refl x) = c x

>>>>

\section{The World of Dependent Types}
In this section, we introduce the concept of dependent types and their role in
type theoretic systems. At their core, dependent types can be thought of simply
as an \textit{interplay of types and terms}. To understand this relationship,
it helps to think of the ways in which types and terms can \textit{depend} on
one another. Altogether, this results in four dependencies:

@@

\(\Pi\)-types are actually significantly more expressive than the ones in
\(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}).

%% In a later system, we
%% allow \(\Pi\)-types to be parameterized over terms (e.g. {\tt ∀ (x : t) → t}).

@@

In the process of introducing dependent types, we have shown the necessary
changes and extensions on the STLC to achieve the expressivity necessary to
function as a foundation for mathematics. To clarify this process and to show
the relationship between each of the systems we have previously mentioned, we
include a diagram of the {\em Lambda Cube} \cite{}:
\begin{center}
  \begin{tikzpicture}[
      back line/.style={densely dotted},
      cross line/.style={preaction={draw=white, -,
          line width=6pt}}]
    \matrix (m) [matrix of math nodes,
      row sep=1em, column sep=.75em,
      text height=1ex,
      text depth=0.5ex]{
      & \lambda\omega & & \lambda C \\
      \lambda 2 & & \lambda P2 \\
      & \lambda\underline\omega & & \lambda P\underline\omega \\
      \lambda\!\!\rightarrow  & & \lambda P \\
    };
    \path[-]
    (m-1-2) edge (m-1-4)
    edge (m-2-1)
    edge [back line] (m-3-2)
    (m-1-4) edge (m-3-4)
    edge (m-2-3)
    (m-2-1) edge [cross line] (m-2-3)
    edge (m-4-1)
    (m-3-2) edge [back line] (m-3-4)
    edge [back line] (m-4-1)
    (m-4-1) edge (m-4-3)
    (m-3-4) edge (m-4-3)
    (m-2-3) edge [cross line] (m-4-3);
  \end{tikzpicture}
\end{center}
Reading the cube requires us to start on the edge marked \(\lambda\!\!\rightarrow\),
which corresponds to the STLC, the base system. From here, we can either move
inwards, upwards, or rightwards, which corresponds to the extension to the other
systems marked on the respective edges. Using these three movements, we
eventually reach \(\lambda C\), leaving it in the most powerful position
on the cube.

@@

prop' : {A : ⊥ → Set} → (x : ⊥) → A x
prop' {A} perp = ind⊥ A perp

absurd : ⊥ → (1 ≡ 2)
absurd = rec⊥ (1 ≡ 2) 
\end{code}
\end{minipage}
\end{center}
The behavior of the empty type is meant to model {\it ex-falso} (i.e., from
absurdity comes anything we like). In general, we use the empty type when
writing proofs containing negated expressions.

@@

ind⊤ : (C : ⊤ → Set) → C * → (x : ⊤) → C x
ind⊤ C c * = c
\end{code}
\end{minipage}
\end{center}
Similar to \(\perp\), \(\top\) is simple and is rather uninteresting by itself.
There are, however, some trivial yet interesting proofs that involve the two.
For the first of these, we also introduce a special kind of proof: {\it uniqueness
of identity} (UIP) \cite{}. In this proof, we set out to determine and classify
the identity of every element within a given type. For example, for \(\top\), we
should be able to prove:
$$\forall (x : \top) \rightarrow x ≡ *$$
That is, for every element of type \(\top\), that element is equivalent to {\tt *}.
This second expression, {\it x ≡} {\tt *}, is an example of an {\it identity type},
which we describe later on in this section. For now, we simply say that the
sole constructor for the identity type is {\tt (refl x)}, which is a proof that
{\it x ≡ x}.

Now, to prove the above proposition, we can use the induction principle for
\(\top\), since the proposition we are trying to prove concerns all elements
of \(\top\):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ : (x : ⊤) → x ≡ *
uip⊤ = ind⊤ (λ top → top ≡ *) (refl *)
\end{code}
\end{minipage}
\end{center}
Here, we have taken our motive to be the expression {\tt (λ top → top ≡ *)},
which essentially repeats the information already given by our type expression.
The second element, {\tt (refl *)}, is a proof that our motive holds in the case that
{\tt top} is {\tt *} and is of type {\tt * ≡ *}.

Alternatively, we could have written our proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ * = refl *
\end{code}
\end{minipage}
\end{center}
This form of proof uses pattern matching and asserts that the only element in
\(\top\) is {\tt *}, which greatly simplifies our proof. Both methods of proof
are correct, but, in general, using pattern matching is only safe when the
values of proof expressions are not dependent one another (i.e., their values
don't change depending on which expression we apply pattern matching to). This
topic is formally known as {\it dependent pattern-matching}, which is a bit
out of the scope of this paper. Thus, for our purposes, we refrain from using
pattern matching in our proofs and instead use recursion and induction
principles.

@@

\section{Equivalence and Sample Proofs}
In this section, we present the type of Natural numbers, Lists, and the identity
type. To further understand these types, we also include a brief description of
the different forms of equality, namely definitional, judgmental and
propositional. In type theory, the specific differences between the three are
generally misunderstood but are important in understanding basic type theoretic
proof methods, which we also go into some detail in this section with several
examples.

\subsection*{Equality}
%% talk about definitional equality, judgmental equality, prop equality
The first form of equality that is generally more familiar than the others.
This form of equality is called {\it definitional equality} \cite{}. Simply put,
definitional equality is equality through definition, which requires no proof.
For example, we know that for any two natural numbers {\tt m} and {\tt n},
{\tt m `plus` n} is equal to the sum of {\tt m} and {\tt n} by definition of
{\tt `plus`}.

We have already briefly mentoined the second form of equality, judgmental
equality \cite{}, which is equality by logical bijection. For example, if we
have the propositions {\tt A} and {\tt B}, and from either {\tt A} or {\tt B} we
can derive {\tt B} and {\tt A} respectively, then we can say that {\tt A} and
{\tt B} are judgmentally equivalent. A proof of judgmental equality requires a
proof similar to the one we have presented in the previous section for the
proposition {\tt (⊤ + ⊤) ↔ 𝔹}.

The final form of equivalence is quite a bit more complex than the previous two
and requires a special kind of proof. Propositional equality is equality through
an identity type (e.g. {\tt x ≡ y}). Thus, we now formally introduce the
definition of the identity type:
%% say something that this is the HoTT version of identity
%% possibly also point out that the discrepancy between the two
%% -- "PROPOSITIONAL EQUALITY"
%% -- For a proposition, C, on a proof of x ≡ y for
%% -- any x and y, if we can show that it holds for
%% -- (refl x), then C holds for all proofs, p : x ≡ y.
%% -- "STRICT EQUALITY"
%% -- For a proposition, C, on a reflexive equality (x ≡ x),
%% -- for any x, if we can show that it holds for (refl x), then
%% -- C holds for all other reflexive equalities p : x ≡ x.
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _≡_ {A : Set} : (a b : A) → Set where
  refl : (a : A) → (a ≡ a)
\end{code}
\end{minipage}
\end{center}
Out of all the types we have introduced, the identity type is the first type
parameterized by terms (viz. {\tt a} and {\tt b}). Until now, we have only
introduced types parameterized by other types. Types paramterized by terms are
classified as an {\em indexed type} \cite{}. Essentially, if we think of types
as shapes (i.e., we can think of them as circles containing points representing
their terms), indexed types change their shape depending on which terms are
passed to them. For example, it is not the case that {\tt y ≡ x} automatically
comes from a proof of {\tt x ≡ y}. This requires a proof!

Before we can prove this, however, we must first introduce the recursion and
induction principles of the identity type. Unlike the other types we have
previously introduced, the identity type has two induction principles, both of
which have their own appropriate use case \cite{}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
J : {A : Set} (C : {x y : A} → x ≡ y → Set) →
    (∀ (x : A) → C (refl x)) →
    {x y : A} (p : x ≡ y) → C p
J C c (refl x) = c x

K : {A : Set} (C : {x : A} → x ≡ x → Set) →
    (∀ (x : A) → C (refl x)) →
    {x : A} (p : x ≡ x) → C p
K C c (refl x) = c x

%% \subsection*{Booleans}
%% \begin{center}
%% \begin{minipage}{0.9\textwidth}
%% \begin{code}
%% data 𝔹 : Set where
%%   True  : 𝔹
%%   False : 𝔹

%% rec𝔹 : (C : Set) → C → C → 𝔹 → C
%% rec𝔹 C t f True = t
%% rec𝔹 C t f False = f

%% ind𝔹 : (C : 𝔹 → Set) → C True → C False → (x : 𝔹) → C x
%% ind𝔹 C t f True = t
%% ind𝔹 C t f False = f
%% \end{code}
%% \end{minipage}
%% \end{center}

@@
